Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-03-18 manuscript/manuscript.tex
manuscript updated

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks with higher association scores to this phenotype. We interrogate a familial breast cancer genome-wide association study focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$ ) and occupy more central positions in the network than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. 
\end{abstract}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with large and moderate effects can be detected. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically Bonferroni. However, Bonferroni is known to be overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

In this regard many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and between them are the edges that link them. These functional relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011} \cite{pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Additionally, as genes that contribute to a disease tend to participate in similar biological functions, studying the neighborhood of known disease genes is effective at identifying new ones \cite{huang_systematic_2018}.

Network-based, biomarker discovery methods exploit this relatedness to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on \emph{star} subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on larger subnetworks interconnected genes with high association scores. However, they differ in their tolerance to the inclusion of low scoring nodes, and the possible number of disconnected subnetworks in the solution. Lastly, other methods also consider the topology of the network, favoring solutions that are not only high scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy - Weinberg equilibrium in controls (P-value \textless~0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes. After control for relatedness, 17 additional samples were removed (6 for sample identity error, 6 controls related to other samples, and 3 additional controls having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with CLIS, 1 with a \emph{BRCA1/2} mutation detected in the family, 1 with unknown molecular diagnosis). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

@@

We use such mapping to compare the outputs of methods that produce SNP-lists to those that produce gene-lists, and vice versa. For the former, we consider any gene that can be mapped to any of the selected SNPs as selected as well. For the latter, we consider all the SNPs that can be mapped to that gene as selected by the method.

@@

\subsubsection{Networks used}
\label{methods:networks}

\paragraph{Gene-gene network}
Out of the six methods tested, five use a gene-gene interaction network (Section~\ref{methods:methods}). Although their respective statistical frameworks are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.), for practical reasons we focused on a protein-protein interaction network (PPIN), as they are interpretable, well characterized, and most of the methods were designed to scale appropriately to it. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless specified otherwise, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the network. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

@@

\paragraph{SNP networks}
SConES \cite{azencott_efficient_2013} is the only of the studied methods designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. \citet{azencott_efficient_2013} proposed three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists, on top of GM (``GI network''). We focused on the GI network, as it fits the scope of this article better, using the PPIN described above. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section~\ref{methods:node_score}. In all three the node scores are the association score of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}).
\subsubsection{Methods used}
\label{methods:methods}

@@

Genes that contribute to the same function are nearby in the PPIN, might be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). How to account for that is not the only choice a developer of network methods needs to make: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, which network to use, etc. In consequence, multiple solutions have been proposed. In this article, we examine six of them: five that explore the PPIN, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table~\ref{tab:method_comparison}.

@@

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subgraph on the network \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P-value into a score that is negative under no association with the phenotype, and positive value when there is. This transformation is achieved by modelling the distribution of P-values by a beta-uniform model (BUM) parameterized by the desired FDR. Thus formulated, the problem is NP-complete, and hence solving it would require a prohibitely long computational time. To solve it efficiently it is re-casted as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S):

@@

were p(v)~=~w(v) - w' is the \emph{profit} of adding a node, c(e)~=~w' is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$. All three are positive quantities. Heinz implements the algorithm from \citet{ljubic_algorithmic_2006}, which although in practice is often fast and optimal, neither is guaranteed. We used BioNet's implementation of heinz, available on Bioconductor \cite{beisser_bionet:_2010,heinz}.

@@

\item[{LEAN}] LEAN searches disregulated ``star'' gene subnetworks, that is, subnetworks composed by one central node and all its interactors \cite{gwinner_network-based_2016}. By imposing this restriction, LEAN is able to exhaustively test all such subnetworks (one per node). For a particular subnetwork of size \emph{m}, the P-values corresponding to the involved nodes are ranked as p\textsubscript{1} \(\le\) \dots{} \(\le\) p\textsubscript{m}. Then, \emph{k} binomial tests are conducted, to compute the probability of having \emph{k} out of \emph{m} P-values lower or equal to p\textsubscript{k} under the null hypothesis. The minimum of these \emph{k} P-values is the score of the subnetwork. This score is transformed into a P-value through an empirical distribution obtained via a subsampling scheme, where sets of \emph{m} genes are selected randomly, and their score computed. Lastly, P-values are corrected for multiple testing through a Benjamini-Hochberg correction. We used the implementation of LEAN from the LEANR R package \cite{leanr}.
\item[{SConES}] SConES searches the minimal, modular, and maximally associated subnetwork in a SNP graph \cite{azencott_efficient_2013}. Specifically, it solves the problem

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. Given two hyperparameters, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. We used the version on SConES implemented in the R package martini \cite{martini}. As in \citet{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. Note that the solution to the above problem can consist of several connected subnetworks which are disconnected from each other. In this case, the selected hyperparameters were \(\eta\)~=~3.51, \(\lambda\)~=~210.29 for SConES GS; \(\eta\)~=~3.51, \(\lambda\)~=~97.61 for SConES GM; and \(\eta\)~=~3.51, \(\lambda\)~=~45.31 for SConES GI.
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the aforementioned methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). Pathways with an adjusted P-value \textless 0.05 were regarded significant.

\subsection{Evaluation of methods}

@@

\subsubsection{Properties ot the solution}
\label{methods:comparison}

We compared the runtime, the number of selected genes/SNPs, and the stability (sensitivity of the result to small changes in the input) of the different network methods. The stability was measured using the Pearson's correlation between different runs as suggested \citet{nogueira_measuring_2016}.

\subsubsection{Classification accuracy of selected biomarkers}
\label{methods:classifier}

A desirable solution offers good predictive power on unseen \emph{test} samples. We evaluated the predicting power of the SNPs selected by the different methods through the performance of an L1-penalized logistic regression classifier trained exclusively on those SNPs to predict the outcome (case/control). The L1 penalty helps to account for LD by reducing the number of SNPs included in the model (size of the active set), while improving the generalization of the classifier. This penalty was set by cross-validation. We applied each network method to each \emph{train} set, and trained the classifier on the \emph{train} set as well using only on the selected SNPs. When the method retrieved a list of genes (all of them except SConES), we considered as selected all the SNPs mapped to any of those genes. Then we evaluated the sensitivity and the specificity on the \emph{test set}. To obtain a baseline, we also trained the classifier on all the SNPs. We also evaluated the size of the active set, an estimate of a plausible, more sparse solution with a comparable predictive power to the original solution.

\subsubsection{Comparison to state-of-the-art}
\label{methods:bcac}
An alternative way to evaluate the results is comparing our results to an external dataset. For that purpose, we recovered a list of 153 genes associated to familial breast cancer from DisGeNET \cite{pinero_disgenet:_2017}. Across this article we refer to these genes as \emph{breast cancer susceptibility genes}.

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC), a meta-analysis of case-control studies conducted in multiple countries which included 228\,951 women of European ancestry from the general population \cite{Michailidou2017}. Hence, a high proportion of breast cancer cases investigated in BCAC are sporadic, while GENESIS is a homogeneous dataset focusing on the French high-risk population attending the family cancer clinics. Despite these differences, there should be shared genetic architecture. On top of that, that overlap should become larger at the gene level. For that purpose, we searched for associated genes as in Section~\ref{methods:node_score}. After imputation, it consisted of 13\,250\,642 SNPs. We used all of them to measure gene-level association, but only the SNPs available on GENESIS after quality control for the SNP-level comparisons.

@@

This work required developing computational pipelines for several GWAS analyses, such physically mapping SNPs to genes, computing gene scores, and performing six different network analyses. For each of those processes, a streamlined, project-agnostic pipeline with a clear interface was created. They are compiled in the following GitHub repository: \url{https://github.com/hclimente/gwas-tools}. The code that applies these pipelines to the GENESIS project, as well as the code that reproduces all the analyses in this article are available at \url{https://github.com/hclimente/genewa}. Although the GENESIS dataset is not publicly available, the published code should work on any other GWAS dataset. All the produced gene subnetworks were deposited on NDEx (\url{http://www.ndexbio.org}), under the UUID e9b0e22a-e9b0-11e9-bb65-0ac135e8bacf.

@@

We conducted association analyses in the GENESIS dataset at both the SNP and the gene levels (Section~\ref{methods:node_score}). Two genomic regions have a P-value lower than the Bonferroni threshold in chromosomes 10 and 16 (Supplementary Figure~\ref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} were related to breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1/2} carriers studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for \emph{BRCA1/2} \cite{rinella_genetic_2013}. Only the peak in chromosome 10 replicated in the gene-level analysis, with \emph{FGFR2} just above the threshold of significance (Supplementary Figure~\ref{sfig:snp_gene_manhattan}B). These results show an overlap in the genetic architecture of the disease between the studied French population sample and other populations, especially at the gene level.

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P-values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. Exploring algorithms unconstrained by considerations about statistical significance we applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}), a machine learning algorithm that searches a small subset of SNPs which provide good classification accuracy. It selected 100 SNPs, from all the aforementioned regions and others (Supplementary Figure~\ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P-value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what the omnigenic model predicts, the genes to which this SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparable classification accuracies to L1-penalized logistic regression (Figure~\ref{fig:benchmark}D), while providing more interpretable solutions.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[height=.75\textheight]{./figures/figure_1.png}
  \caption{\label{fig:solution_overview} Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. As LEAN did not produce any significant gene (adjusted P-value < 0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)} Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54 \texttimes{} 10\textsuperscript{-7} for SNPs, 1.53 \texttimes{} 10\textsuperscript{-6} for genes). \textbf{(B)} Solution networks produced by the different methods. \textbf{(C)} Pearson correlation between the genes selected by each of the methods. \textbf{(D)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs/genes in GENESIS (left) and BCAC (middle), and breast cancer susceptibility genes (right, only at the gene-level). For reference, we added a gray line with a slope of 1.}
\end{figure}

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (adjusted P-value < 0.05), we obtained six solutions (Figure~\ref{fig:solution_overview}, Supplementary~Files~1~and~2): one for each of the remaining four gene-based methods, one for SConES GI, which works at the SNP level, and the consensus.

>>>>

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks with higher association scores to this phenotype. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$ ) and occupy more central positions in the network than average. Additionally, it shows a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Finally, we find a significantly large overlap between the genes in the consensus network and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions.
\end{abstract}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the aforementioned Bonferroni. However, Bonferroni is overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

In this regard many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Additionally, as genes that contribute to a disease tend to participate in similar biological functions, studying the neighborhood of known disease genes is effective at identifying new ones \cite{huang_systematic_2018}.

Network-based, biomarker discovery methods exploit this relatedness to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on larger subnetworks interconnected genes with high association scores. However, they differ in their tolerance to the inclusion of low scoring nodes, and the possible number of disconnected subnetworks in the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy - Weinberg equilibrium in controls (P-value \textless~0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes. After controlling for relatedness, 17 additional samples were removed (6 for sample identity error, 6 controls related to other samples, and 3 additional controls having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with lobular carcinoma \emph{in situ}, 1 with a \emph{BRCA1/2} mutation detected in the family, 1 with unknown molecular diagnosis). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

@@

We used such mapping to compare the outputs of methods that produce SNP-lists to those that produce gene-lists, and vice versa. For the former, we considered any gene that can be mapped to any of the selected SNPs as selected as well. For the latter, we considered all the SNPs that can be mapped to that gene as selected by the method.

@@

\subsubsection{Networks}
\label{methods:networks}

\paragraph{Gene network}
Out of the six methods tested, five use a gene-gene interaction network (Section~\ref{methods:methods}). Although their respective statistical frameworks are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.), we used protein-protein interaction networks (PPIN), as they are interpretable, well characterized, and most of the methods were designed to scale appropriately to it. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless specified otherwise, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the network. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

@@

\paragraph{SNP networks}
SConES \cite{azencott_efficient_2013} is the only studied methods designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. \citet{azencott_efficient_2013} proposed three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists, on top of GM (``GI network''). We focused on the GI network, as it fits the scope of this work better, using the PPIN described above. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section~\ref{methods:node_score}. In all three the node scores are the association scores of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}).

\subsubsection{Studied methods}
\label{methods:methods}

@@

Genes that contribute to the same function are nearby in the PPIN, might be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). But accounting for that is not the only choice a developer of a network method needs to make: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, which network to use, etc. Unsurprisingly, multiple solutions have been proposed. We examined six of them: five that explore the PPIN, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table~\ref{tab:method_comparison}.

@@

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subgraph on the network \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P-value into a score that is negative under no association with the phenotype, and positive when there is. This transformation is achieved by modelling the distribution of P-values by a beta-uniform model (BUM) parameterized by the desired FDR. Thus formulated, the problem is NP-complete, and hence solving it would require a prohibitely long computational time. To solve it efficiently it is re-casted as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S):

@@

were p(v)~=~w(v) - w' is the \emph{profit} of adding a node, c(e)~=~w' is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$. All three are positive quantities. Heinz implements the algorithm from \citet{ljubic_algorithmic_2006} which, although in practice is often fast and optimal, neither is guaranteed. We used BioNet's implementation of heinz \cite{beisser_bionet:_2010,heinz}.

@@

\item[{LEAN}] LEAN searches altered ``star'' subnetworks, that is, subnetworks composed by one central node and all its interactors \cite{gwinner_network-based_2016}. By imposing this restriction, LEAN is able to exhaustively test all such subnetworks (one per node). For a particular subnetwork of size \emph{m}, the P-values corresponding to the involved nodes are ranked as p\textsubscript{1} \(\le\) \dots{} \(\le\) p\textsubscript{m}. Then, \emph{k} binomial tests are conducted, to compute the probability of having \emph{k} out of \emph{m} P-values lower or equal to p\textsubscript{k} under the null hypothesis. The minimum of these \emph{k} P-values is the score of the subnetwork. This score is transformed into a P-value through an empirical distribution obtained via a subsampling scheme, where gene sets of the same size are selected randomly, and their score computed. Lastly, P-values are corrected for multiple testing through a Benjamini-Hochberg correction. We used the implementation of LEAN from the LEANR R package \cite{leanr}.
\item[{SConES}] SConES searches the minimal, modular, and maximally associated subnetwork in a SNP graph \cite{azencott_efficient_2013}. Specifically, it solves the problem

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. Given a $\lambda$ and an $\eta$, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. As in \citet{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. In this case, the selected hyperparameters were \(\eta\)~=~3.51, \(\lambda\)~=~210.29 for SConES GS; \(\eta\)~=~3.51, \(\lambda\)~=~97.61 for SConES GM; and \(\eta\)~=~3.51, \(\lambda\)~=~45.31 for SConES GI. We used the version on SConES implemented in the R package martini \cite{martini}.
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the aforementioned methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). Pathways with an adjusted P-value \textless~0.05 were deemed significant.

\subsection{Evaluation of methods}
\label{methods:comparison}

@@

\subsubsection{Properties ot the solution}
\label{methods:algorithm_comparison}

We compared the runtime, the number of selected genes/SNPs, and the stability (sensitivity of the result to small changes in the input) of the different network methods. The stability was measured using the Pearson's correlation between different runs as suggested by \citet{nogueira_measuring_2016}.

\subsubsection{Classification accuracy of selected SNPs}
\label{methods:classifier}

A desirable solution offers good predictive power on unseen \emph{test} samples. We evaluated the predicting power of the SNPs selected by the different methods through the performance of an L1-penalized logistic regression classifier trained exclusively on those SNPs to predict the outcome (case/control). The L1 penalty helps to account for LD by reducing the number of SNPs included in the model (size of the active set), while improving the generalization of the classifier. This penalty was set by cross-validation. We applied each network method to each \emph{train} set, and trained the classifier on it as well using only on the selected SNPs. When the method retrieved a list of genes (all of them except SConES), we considered as selected all the SNPs mapped to any of those genes. Then we evaluated the sensitivity and the specificity on the \emph{test set}. The active set gave an estimate of a plausible, more sparse solution with a comparable predictive power to the original solution. To obtain a baseline, we also trained the classifier on all the SNPs.

\subsubsection{Comparison to state-of-the-art}
\label{methods:bcac}

An alternative way to evaluate the results is comparing our results to an external dataset. For that purpose, we recovered a list of 153 genes associated to familial breast cancer from DisGeNET \cite{pinero_disgenet:_2017}. Across this article we refer to these genes as \emph{breast cancer susceptibility genes}.

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC), a meta-analysis of case-control studies conducted in multiple countries which included 13\,250\,642 SNPs (after imputation) genotyped on 228\,951 women of European ancestry from the general population \cite{Michailidou2017}. Hence, a high proportion of breast cancer cases investigated in BCAC are sporadic, while GENESIS is a homogeneous dataset focusing on the French high-risk population attending the family cancer clinics. Despite these differences, there should be shared genetic architecture, specially at the gene level. For that purpose, we searched for associated genes as in Section~\ref{methods:node_score}. We used all the SNPs to measure gene association, but only those available after quality control on GENESIS for the SNP-level comparisons.

@@

We developed computational pipelines for several steps of GWAS analyses, such as physically mapping SNPs to genes, computing gene scores, and performing six different network analyses. For each of those processes, we created a pipeline with a clear interface that should woth on any GWAS dataset. They are compiled in \url{https://github.com/hclimente/gwas-tools}. Although the GENESIS data is not public, the code that applies these pipelines to it, as well as the code that reproduces all the analyses in this article are available at \url{https://github.com/hclimente/genewa}. We deposited all the produced gene subnetworks on NDEx (\url{http://www.ndexbio.org}), under the UUID e9b0e22a-e9b0-11e9-bb65-0ac135e8bacf.

@@

We conducted association analyses in the GENESIS dataset at both SNP and gene levels (Section~\ref{methods:node_score}). Two genomic regions have a P-value lower than the Bonferroni threshold in chromosomes 10 and 16 (Supplementary Figure~\ref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} were related to breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1/2} carriers studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for \emph{BRCA1/2} \cite{rinella_genetic_2013}. Only \emph{FGFR2} was significant at the gene-level (Supplementary Figure~\ref{sfig:snp_gene_manhattan}B). These results show an overlap in the genetic architecture of the disease between the studied French population sample and other populations.

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P-values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}), a machine learning algorithm that searches a small subset of SNPs which provide good classification accuracy. It selected 100 SNPs, from all the aforementioned regions and others (Supplementary Figure~\ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P-value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what the omnigenic model predicts, the genes to which this SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparable classification accuracies to L1-penalized logistic regression (Figure~\ref{fig:benchmark}D), while providing more interpretable solutions.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[height=.75\textheight]{./figures/figure_1.png}
  \caption{\label{fig:solution_overview} Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. As LEAN did not produce any significant gene (adjusted P-value < 0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)} Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54 \texttimes{} 10\textsuperscript{-7} for SNPs, 1.53 \texttimes{} 10\textsuperscript{-6} for genes). \textbf{(B)} Pearson correlation between the genes selected by each of the methods. \textbf{(C)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs/genes in GENESIS (left) and BCAC (middle), and breast cancer susceptibility genes (right, only at the gene-level). For reference, we added a gray line with a slope of 1. \textbf{(D)} Solution networks produced by the different methods.}
\end{figure}
